---
title: "VPC Demo"
author: "Samuel P Callisto, PhD"
date: "27 August 2021"
output: pdf_document
---

Packages you may not be familiar with:
 - vpc
 - glue
 - here
 - data.table (preferred over read_csv, also works with NONMEM table files)

```{r}
## Required packages
suppressPackageStartupMessages({
  library(tidyverse)
  library(mrgsolve)
  library(vpc)
  library(glue)
  library(here)
  library(data.table)
})

## making plots pretty
mrg_vpc_theme = new_vpc_theme(list(
  sim_pi_fill = "steelblue3", sim_pi_alpha = 0.5,
  sim_median_fill = "grey60", sim_median_alpha = 0.5
))

## programmatically set run number
## easy to run for multiple models this way
runno <- "r2"

## control sequence
writeplots <- TRUE
```

Load in the NONMEM data set, which we will use as a template for our 
VPC simulations

**Question:** What are we trying to determine by doing a VPC?

```{r, include=FALSE, eval=FALSE, echo=FALSE}
Answer: We want to see if the model produces data which matches the data which 
we have observed and built the model using.
```

```{r}
data <- data.table::fread(here("data","dat1.csv"), na=".") %>% 
  ## add columns necessary for mrgsim
  mutate(
    evid=mdv,
    DOSE=ifelse(amt==0,NA_real_,amt),
    cmt=1
  ) %>%
  ## fill DOSE within a subject
  group_by(ID) %>% 
  fill(DOSE, .direction="downup") %>% 
  ungroup()
head(data)
```


```{r}
#' # Simulate the vpc
#' 
#' ## Load the mrgsolve model
#' 
#' This should reflect `../model/nonmem/r2.ctl`
mod <- mread(glue("../model/sim/{runno}.cpp"))

#' # Set up the simulation
#' 
#' Create a function to simulate out one replicate
sim <- function(rep, data, model) {
  mrgsim(
    model, 
    data = data,
    carry_out = "DOSE,evid,NUM",
    Req = "Y", 
    output = "df", 
    quiet  = TRUE
  ) %>%  mutate(irep = rep)
}

#' Simluate data

#' 200 replicates
isim <- seq(200)

set.seed(86486)
sims <- lapply(
  isim, sim, 
  data = data, 
  mod = mod
) %>% 
  bind_rows() 

sum(sims$Y)

#' Filter both the observed and simulated data
fdata <-  rename(data, id=ID) %>% 
  ## remove dosing rows
  filter(evid == 0)
fsims <- rename(sims, id=ID) %>% 
  ## remove dosing rows
  filter(evid == 0)
```

# Creating the VPC plot

The `vpc` function from Ron Keizer's `vpc` package does most of the heavy 
lifting here. The documentation and vignettes on this package are sparse, but 
it is beneficial to spend some time reading the help file for the `vpc` function 
to see all the different argument options.



```{r}
#' # Create the plot
#' 
#' Pass observed and simulated data into vpc function
p1 <- vpc(
  obs = fdata,
  sim = fsims,
  obs_cols = list(dv = "dv"),
  sim_cols=list(dv="Y", sim="irep"), 
  log_y = TRUE,
  pi = c(0.05, 0.95),
  ci = c(0.025, 0.975), 
  show = list(obs_dv = TRUE), 
  vpc_theme = mrg_vpc_theme
) +  
  ## use ggplot to alter aesthetics
  theme_bw() + 
  xlab("Time (hours)") + 
  ylab("Drugx concentration (mg/mL)")

p1

## saves to png/pdf
if(writeplots){
  mrggsave::mrggsave(p1, stem = "pk-vpc-{runno}", height = 7.5,
                     dir = here("deliv","figure"), script = "pk-vpc-final.R")
}
```

As the analyst we have a couple of options with our VPC:
 - stratify by dose (or other relevant covariate!)
 - dose-normalized (e.g. `DVN = DV / DOSE`)
 - prediction-corrected VPC
 
**Question:** What would be a case when we would need to consider these options?

```{r, include=FALSE, eval=FALSE, echo=FALSE}
Answer: we want to see if the model will produce data which matches that which 
we built the model using. If the VPC produces different trends, it tells us that 
the model is not accurately describing the data. We can hone in specifically on 
which parameter isn't working (e.g. terminal elimination phase, absorption 
phase, missing Cmax implies volume may be off, etc.)
```

**Question:** How could we determine if the model will accurately predict 
future data?

```{r, include=FALSE, eval=FALSE, echo=FALSE}
Answer: An approach commonly done to answer this is an external validation, in 
which a subset of data is withheld, then we can overlay it on the VPC to see how 
the model produces data compared to FUTURE OBSERVED data. Notice how this is 
answering a different question than the typical VPC.
```

## Pred-corrected VPC

Typically we only need to do this if we have different doses in the data set. 
But I will just do one here so you can see the required input

Need to get PREDs. Two options:
 - can pull from NONMEM table file
 - can simulate from model using NONMEM

```{r}
tab <- data.table::fread(
  here("model","nonmem",runno,glue("{runno}.tab")), skip=1
)
head(tab)
```

Key reference: Bergstrand, 2011
https://pubmed.ncbi.nlm.nih.gov/21302010/

Update data sets to include PREDs
```{r}
#' Now, remove random effects and simulate PREDs
pred <- sim( 
  rep=1,
  data = data, 
  mod = zero_re(mod)
)
```

## compare NONMEM and mrgsolve PREDs

They match entirely, so it doesn't matter which method you use. PREDs are 
deterministic since they zero-out the random effects.

```{r}
full_join(
  rename(tab, TABPRED=PRED), 
  rename(pred, NMPRED=Y)
) %>% 
  ggplot() + 
  geom_point(aes(x=NMPRED,y=TABPRED)) + 
  geom_abline(intercept=0, slope=1)
```


```{r}
## modify data to include obs only and append PREDs
fdata <- data %>% 
  rename(id=ID) %>% 
  full_join(
    select(pred, NUM, PRED=Y)
  ) %>% filter(evid == 0)

fsims <-  sims %>% 
  rename(id=ID) %>% 
  full_join(
    select(pred, NUM, PRED=Y)
  ) %>% filter(evid == 0)

#' # Create the plot
#' 
#' Pass observed and simulated data into vpc function
pc1 <- vpc(
  obs = fdata,
  sim = fsims,
  pred_corr = TRUE,
  obs_cols = list(dv = "DV", idv="time", pred="PRED"),
  sim_cols=list(dv="Y", idv="time", pred="PRED", sim="irep"),
  log_y = TRUE,
  pi = c(0.05, 0.95),
  ci = c(0.025, 0.975), 
  vpc_theme = mrg_vpc_theme
) +
    ## use ggplot to alter aesthetics
  theme_bw() + 
  xlab("Time (hours)") + 
  ylab("Prediction-corrected concentration (mg/mL)")

pc1

## write plots
if(writeplots){
  mrggsave::mrggsave(p1, stem = "pk-vpc-{runno}-pred-corr", height = 7.5,
                     dir = here("deliv","figure"), script = "pk-vpc-final.R")
}
```

